The Kapowski Chronicles: A Complete, Volumetric Cortical Thickness Open Source Pipeline with Evaluation on Public Data

### 2013-05-08 KapowskiChronicles/methods.tex
ENH:  Amy

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternating between prior-based segmentation and white matter posterior
        probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

%We also note that each component is publicly available with all ANTs 
%algorithms available as open source.%
%\footnote{
%http://www.picsl.upenn.edu/ANTS
%}
Additionally, the coordination of all the algorithmic components is
encapsulated in the shell script \verb#antsCorticalThickness.sh#.  This includes
optimal parameters for each of the algorithmic components which has worked
well for our processing and which are used to acquire the results 
described in this work.

@@

\begin{figure*}
  \centering
  \includegraphics[width=170mm]{Figures/Kapowski_pipeline2.pdf}
  \caption{The ANTs T1 processing workflow containing all elements for 
  determining cortical thickness. Not shown is the optional single subject
  to template registration.}
  \label{fig:pipeline}
\end{figure*}

@@

Normalizing images to a standard coordinate system
reduces intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as the selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \citep{Talairach1988}, or a publicly available averaged group of
subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
templates.  Additionally, mean templates constructed from labeled
data can be used to construct spatial priors for improving segmentation
algorithms.
The work of \cite{avants2010} explicitly models the geometric component of the 
normalized space during optimization to produce such mean templates.  Coupling the intrinsic symmetry of 
SyN pairwise registration \citep{avants2011} and an
optimized shape-based sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \citep{avants2010} with arbitrary similarity metric choice.

@@

The ANTs implementation of this technique is currently available as a shell script, 
\verb#buildtemplateparallel.sh#, and a multivariate version,
\verb#antsMultivariateTemplateConstruction.sh#, both of which are distributed as part of
 the ANTs repository.
The latter script permits the construction of multimodal templates (e.g. 
T1-weighted, T2-weighted, and proton density MRI as described in the 
Evaluation section).  Both scripts accommodate a variety of computational resources
for facilitating template construction.  These computational resource possibilities include:
\begin{itemize}
  \item serial processing on a single workstation, 
  \item parallelized processing on a single workstation with multiple cores using \verb#pexec#%
  \footnote{http://www.gnu.org/software/pexec/pexec.1.html},
  \item parallelized processing using Apple's XGrid technology%
  \footnote{https://developer.apple.com/hardwaredrivers/hpc/xgrid\_intro.html}, 
  \item parallelized processing using Sun Grid Engine for cluster-based systems%
  \footnote{http://www.oracle.com/technetwork/oem/grid-engine-166852.html}, and 
  \item parallelized processing using the Portable Batch System for cluster-based systems%
  \footnote{http://www.pbsworks.com/}.
\end{itemize}
Within this work multiple templates were created for all stages of 
image processing and analysis.  The creation of these templates are described
in the corresponding data section.

\textcolor{blue}{what is the motivation of the multivar template here?
  it would seem useful to add an FA component to the segmentation step
  to aid in cortical / wm delineation which would motivate the
  multivar template approach but might be difficult to establish that
  this actually helps .... though we could compare both ways : univar
  and multivar }

@@

In \cite{tustison2010}, we introduced an extension of N3, denoted as
N4, which demonstrates improved performance and convergence behavior
on a variety of data.  This improvement is a result of an enhanced 
fitting routine (which includes multi-resolution capabilities) and a modified optimization 
formulation.  For our workflow, the additional possibility of specifying
a weighted mask in N4 permits the use of the current white matter probability map 
calculated during the segmentation pipeline for further improvement of 
bias field estimation.  In addition to its public availability 
through ANTs and the Insight Toolkit, it has also been included in
the popular open source Slicer software package for visualization and medical
image computing \cite{fedorov2011}.

N4 is used in two places during the individual subject processing (cf. Figure
\ref{fig:pipeline}).  Following conversion of the raw dicom T1-weighted image
to Nifti format using our related \verb#Neuropipedream# set of raw image conversion
and organization tools%
\footnote{
http://sourceforge.net/projects/neuropipedream/
}, N4 is used to generate an initial bias corrected image for use in
brain extraction.  The input mask is created by adaptively thresholding 
the background from the foreground using Otsu's algorithm \cite{otsu1979}.
Following brain extraction, the three-tissue segmentation involves iterating
between bias field correction using the current white matter posterior 
probability as a weight mask and then using that bias corrected image
as input to the Atropos segmentation step (described in subsequent sections).

@@

@@

\subsubsection{DiReCT Cortical Thickness Estimation}

Although the basic formulation of DiReCT as reported in this work is as it 
was introduced in \cite{das2009}, we have made several 
improvements.  Perhaps the most significant advance is that this particular 
ITK-compatible implementation has been significantly multi-threaded,
is written in ITK coding style, and has been made publicly available through 
ANTs complete with a unique user interface design developed specifically for 
ANTs tools.

@@

\subsubsection{LPBA40 Data for Skull Stripping}

For the brain extraction step
we used the data from the LPBA40 repository \citep{shattuck2008}.
These data consist of 40 high-resolution 3D Spoiled Gradient Echo
(SPGR) MRI acquisitions which were manually labeled delineating
56 brain structures.  Additional post processing 
included automated brain extraction using FSL's brain extraction tool 
\citep{smith2002} which was followed by manual corrections.  These
40 brain masks are also included with the database.

All 40 subjects were used to create a population-specific
unbiased average template \cite{avants2010}.  The brain masks corresponding 
to the 20 subjects were warped to the template space using the 
transforms derived during the template building process.  A template 
probability mask was created by averaging the warped brain masks.
For brain extraction of any single individual this ANTS-based LPBA40 template 
is coarsely registered to the single subject brain.  The template
probability brain mask is warped to the individual subject and 
is used as the initial brain mask estimate.  As mentioned previously,
Atropos and binary morphological operations are used to refine the
brain mask estimate.

\subsubsection{NIREP Data for 3-Tissue Segmentation}

The nonrigid image registration evaluation project is an ongoing 
framework for evaluating image registration algorithms \citep{christensen2006}.
The initial data set introduced into the project consists of 
16 (8 male and 8 female) high resolution skull-stripped brain 
data with 32 cortical labels (cf Table \ref{table:nirep_labels}) manually drawn using a 
published protocol.%
\footnote{
http://www.nirep.org/
}
Given the gray matter labels, the white matter and CSF were identified 
for each of the 16 subjects using Atropos.  Similar to the LPBA40
data set, a NIREP template was created from all 16 subjects and each of the
warped labels were used to create probabilistic estimates of the 
labeled region boundaries. These probability maps were used as 
spatial prior probabilities during the 3-tissue segmentation component
of the pipeline.  Using SyN, the NIREP template is registered to the
extracted individual subject brain which is followed by a warping of the 
NIREP priors to the space of the individual subject.  The initial warped 
white matter probability map is used as the weighted confidence mask 
in the follow-up bias correction step.

@@

\subsection{IXI Data for Pipeline Evaluation}

The IXI data%
\footnote{
http://biomedic.doc.ic.ac.uk/brain-development/
}
used for the evaluation consists of 577 healthy subjects imaged at three sites 
using several modalities (T1-weighted, T2-weighted, proton density, magnetic 
resonance angiography, and diffusion tensor imaging).  The 
database also consists of  demographic information such as age, weight,
height, ethnicity, occupation category, educational level, and marital status.
The number of subjects spanning a range of demographic characteristics makes
this a rich data set for validating and exploring correlations with cortical 
thickness measured using the ANTs pipeline.

@@

>>>>

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternation between prior-based segmentation and white matter posterior
        probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

%We also note that each component is publicly available with all ANTs 
%algorithms available as open source.%
%\footnote{
%http://www.picsl.upenn.edu/ANTS
%}
Additionally, the coordination of all the algorithmic components is
encapsulated in the shell scripts \verb#antsCorticalThickness.sh# with
subcomponents delegated to \verb#antsBrainExtraction.sh# 
and \verb#antsAtroposN4.sh#.  This includes
optimal parameters for each of the algorithmic components which has worked
well for our processing and which are used to acquire the results 
described in this work.

@@

\begin{figure*}
  \centering
  \includegraphics[width=170mm]{Figures/Kapowski_pipeline2.pdf}
  \caption{Illustration of the main components of the ANTs processing 
  workflow containing all elements for determining cortical thickness. 
  Not shown is the optional subject to template registration.}
  \label{fig:pipeline}
\end{figure*}

@@

Normalizing images to a standard coordinate system
reduces intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as the selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \citep{Talairach1988}, or a publicly available averaged group of
subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
templates.  Additionally, mean templates constructed from labeled
data can be used to construct spatial priors for improving segmentation
algorithms.
The work of \cite{avants2010} explicitly models the geometric component of the 
normalized space during optimization to produce such mean templates.  Coupling the intrinsic symmetry of 
SyN pairwise registration \citep{avants2011} and an
optimized shape-based sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \citep{avants2010}.

@@

The ANTs implementation of this technique is currently available as a shell script, 
\verb#buildtemplateparallel.sh# (a multivariate version is also available as
\verb#antsMultivariateTemplateConstruction.sh#).  Both scripts are distributed as part of
 the ANTs repository.
The latter script permits the construction of multimodal templates (e.g. 
T1-weighted, T2-weighted, proton density MRI and fractional anisotropy).  
Both scripts accommodate a variety of computational resources
for facilitating template construction.  These computational resource possibilities include:
\begin{itemize}
  \item serial processing on a single workstation, 
  \item parallelized processing on a single workstation with multiple cores using \verb#pexec#%
  \footnote{http://www.gnu.org/software/pexec/pexec.1.html},
  \item parallelized processing using Apple's XGrid technology%
  \footnote{https://developer.apple.com/hardwaredrivers/hpc/xgrid\_intro.html}, 
  \item parallelized processing using Sun Grid Engine for cluster-based systems%
  \footnote{http://www.oracle.com/technetwork/oem/grid-engine-166852.html}, and 
  \item parallelized processing using the Portable Batch System for cluster-based systems%
  \footnote{http://www.pbsworks.com/}.
\end{itemize}

For this work, database-specific templates were used during cortical thickness pipeline
processing for both brain extraction and brain segmentation steps.  Multivariate templates 
were constructed from the multimodal data sets.  However, their usage was based on the 
fact that they had been built previously for other work and not because they provide a 
discernible advantage over univariate templates (i.e. T1-only) for the proposed workflow.

@@

In \cite{tustison2010}, we introduced an improvement of N3, denoted as
``N4'', which demonstrates a significant increase in performance and convergence behavior
on a variety of data.  This improvement is a result of an enhanced 
fitting routine (which includes multi-resolution capabilities) and a modified optimization 
formulation.  For our workflow, the additional possibility of specifying
a weighted mask in N4 permits the use of a ``pure tissue'' probability map 
(described below)
calculated during the segmentation pipeline for further improvement of 
bias field estimation.  In addition to its public availability 
through ANTs and the Insight Toolkit, it has also been included in
the popular open source Slicer software package for visualization and medical
image computing \cite{fedorov2011}.

N4 is used in two places during the individual subject processing (cf. Figure
\ref{fig:pipeline}).  Following conversion of the raw dicom T1-weighted image
to Nifti format using our related \verb#Neuropipedream# set of raw image conversion
and organization tools%
\footnote{
http://sourceforge.net/projects/neuropipedream/
}, N4 is used to generate an initial bias corrected image for use in
brain extraction.  The input mask is created by adaptively thresholding 
the background from the foreground using Otsu's algorithm \cite{otsu1979}.
Following brain extraction, three-tissue segmentation involves iterating
between bias field correction using the current pure tissue 
probability map as a weight mask and then using that bias corrected image
as input to the Atropos segmentation step (described in the next section).

@@

Due to the important interplay between segmentation and bias correction,
we perform multiple N4 $\Leftrightarrow$ Atropos iterations.
In order to better integrate Atropos and N4, we use  
a pure tissue probability weight mask generated from the 
posterior probabilities generated from the segmentation 
process.  Given $N$ labels and the corresponding $N$
posterior probability maps $\{ P_1, \ldots, P_N\}$ produced
during the segmentation process, the N4 weight mask is 
created at each N4 $\Leftrightarrow$ Atropos iteration from
\begin{align}
  P_{pure\,\,tissue}(\mathbf{x}) = \sum_{i=1}^N P_i(\mathbf{x}) \prod_{j=1, j \neq i}^N \left( 1 - P_j(\mathbf{x}) \right).
\end{align}
One of the key insights of the original N3 algorithm is the
observation that inhomogeneities cause the intensity values of
pure tissue peaks to spread in the intensity histogram as though
convolved with a Gaussian.  A core contribution of N3 is the
proposed corrective of deconvolving the intensity histogram to 
accentuate the tissue peaks coupled with a spatial smoothing 
constraint. The pure tissue probability mask
weights more heavily the voxels corresponding to pure tissue 
types (as determined by the segmentation) during the deconvolution process 
while minimizing the contribution of regions such as the gray/white matter 
interface where peak membership is ambiguous.

@@

\subsubsection{DiReCT (aka KellySlater/KellyKapowski) Cortical Thickness Estimation}

DiReCT was introduced 
in \cite{das2009} and made available in ANTs with the program \verb#KellySlater#.
Since then several improvements have been made and incorporated into the program
\verb#KellyKapowski#.%
\footnote{
Traditional academic discourse encountered in the published literature
rarely contextualizes peculiarities such as algorithmic nomenclature.
We briefly mention that
this was the source of a rare disagreement between the first two authors
based, as many disagreements are, on a simple misunderstanding and not an
affronting existential statement concerning a certain favorite sitcom
of the author's youth. 
}
Among the most significant advancements is that the more recent
implementation is multi-threaded, written in rigorous ITK coding style, and 
has been made publicly available through  ANTs complete with a unique user 
interface design developed specifically for ANTs tools.

@@

%\subsubsection{LPBA40 Data for Skull Stripping}
%
%For the brain extraction step
%we used the data from the LPBA40 repository \citep{shattuck2008}.
%These data consist of 40 high-resolution 3D Spoiled Gradient Echo
%(SPGR) MRI acquisitions which were manually labeled delineating
%56 brain structures.  Additional post processing 
%included automated brain extraction using FSL's brain extraction tool 
%\citep{smith2002} which was followed by manual corrections.  These
%40 brain masks are also included with the database.  
%
%All 40 subjects were used to create a population-specific
%unbiased average template \cite{avants2010}.  The brain masks corresponding 
%to the 20 subjects were warped to the template space using the 
%transforms derived during the template building process.  A template 
%probability mask was created by averaging the warped brain masks.
%For brain extraction of any single individual this ANTS-based LPBA40 template 
%is coarsely registered to the single subject brain.  The template
%probability brain mask is warped to the individual subject and 
%is used as the initial brain mask estimate.  As mentioned previously,
%Atropos and binary morphological operations are used to refine the
%brain mask estimate.

\subsubsection{NIREP Data for Cortical Labelings}

The Nonrigid Image Registration Evaluation Project (NIREP%
\footnote{
http://www.nirep.org/
}) 
is an ongoing framework for evaluating image registration algorithms \citep{christensen2006}.
The initial data set introduced into the project consists of 
16 (8 male and 8 female) high resolution skull-stripped brain 
data with 32 cortical labels (cf. Table \ref{table:nirep_labels}) manually drawn using a 
published protocol.

Cortical label propagation to each individual subject for all data sets
described below is performed using the PICSL multi-atlas joint label fusion
algorithm described in \cite{wang2013} which recently topped the competition
at the MICCAI 2012 Grand Challenge on Multi-Atlas Labeling and is also 
distributed with the ANTs toolkit.  .

%Given the gray matter labels, the white matter and CSF were identified 
%for each of the 16 subjects using Atropos.  Similar to the LPBA40
%data set, a NIREP template was created from all 16 subjects and each of the
%warped labels were used to create probabilistic estimates of the 
%labeled region boundaries. These probability maps were used as 
%spatial prior probabilities during the 3-tissue segmentation component
%of the pipeline.  Using SyN, the NIREP template is registered to the
%extracted individual subject brain which is followed by a warping of the 
%NIREP priors to the space of the individual subject.  The initial warped 
%white matter probability map is used as the weighted confidence mask 
%in the follow-up bias correction step.

@@

\subsubsection{Data for Pipeline Evaluation}

\paragraph{IXI} The IXI data%
\footnote{
http://biomedic.doc.ic.ac.uk/brain-development/
}
used for the evaluation consists of 577 healthy subjects imaged at three sites 
using several modalities (T1-weighted, T2-weighted, proton density, magnetic 
resonance angiography, and diffusion tensor imaging).  The 
database also consists of  demographic information such as age, weight,
height, ethnicity, occupation category, educational level, and marital status.
The number of subjects spanning a range of demographic characteristics makes
this a rich data set for validating and exploring correlations with cortical 
thickness measured using the ANTs pipeline.

@@

\paragraph{Oasis}

### 2013-05-08 KapowskiChronicles/paper.tex
ENH: Ozzy
ENH:  Amy

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools
(ANTs) cortical thickness automated pipeline
comprising well-vetted components such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias
correction),
Atropos ($n$-tissue segmentation), and DiReCT (cortical thickness) 
all developed as part of the ANTs open science effort.  
Complementing the open source aspect of ANTs we confirm 
previous findings of
gender differences, age correlation, and hemispherical asymmetry 
in cortical thickness using the publicly available IXI data
consisting of multimodal structural MRI of approximately 600 subjects.
To further promote open science and use of the proposed tools,
 all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep bias correction \sep cortical thickness \sep open source software \sep segmentation \sep skull stripping
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

\input{results}

>>>>

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools (ANTs) cortical thickness automated pipeline
comprising well-vetted components such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias correction), Atropos ($n$-tissue segmentation), 
and DiReCT (cortical thickness) all developed as part of the ANTs open science effort.  
Complementing the open source aspect of ANTs we investigate several hypotheses
explored previously in the literature using four different publicly 
available data sets totaling approximately 1200 subjects.  
To further promote open science and use of the proposed tools,
all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep cortical thickness \sep open science
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

%\input{results}
